---
title: "Explore the dataset"
output:
  html_document:
    df_print: paged
---


```{r results='hide', echo=FALSE, warning=FALSE, message=FALSE}
#Ignore this chunk, it is only to run the setup Rmarkdown when knitting 
building_htmls <- ! interactive()
if (building_htmls ) {
   source(knitr::purl("00_Setup.Rmd", quiet=TRUE))
}
```


## The study

This workshop will use CosMx Spatial Molecular Imager (SMI) data from the paper "Macrophage and neutrophil heterogeneity at single-cell spatial resolution in human inflammatory bowel disease" (Garrido-Trigo et al. 2023) - link: https://www.nature.com/articles/s41467-023-40156-6

There were 9x colon tissue samples, one per slide. They used a 1k RNA panel panel. (5k, 6k and whole transcriptome available) 

* 3x Healthy controls (HC)
* 3x Crohn’s disease (CD)
* 3x Ulcerative colitis (UC)

For this workshop, we will work with a subsetted dataset;

* Only the Healthy and Chrohn's disease samples - 6x samples total
* Only a regional subset of each sample - First 4 'FOVs' (feilds of view - cosmx scans)






## The Data object


This is a 'SpatialFeatureExperiment' Object.


This 'mini' dataset has 999 genes and 65601 cells. 

```{r}
sfe
```



It has the following metadata for each cell.

```{r}
DT::datatable(as.data.frame(head(colData(sfe), n=300)))
```



This was a '1k' panel, we have 999 targets.
```{r}
DT::datatable(as.data.frame(rowData(sfe)))
```



How many cells per sample? 

```{r}
table(sfe$tissue_sample)
```

This data is subsetted to only 4 'Feilds of View' (FOV). On the cosmx platform, these are multiple rectangular regions (feilds of view) that make up the run. Essentially we're looking at a corner of each sample.

```{r}
#NB: ColData is a DataFrame, not a data.frame, often need an explicit conversion
colData(sfe) %>% as.data.frame() %>% select(group,tissue_sample, fov, fov_name) %>%
  group_by(group,tissue_sample, fov, fov_name) %>% 
  summarise(n_cells = n())
```




## The existing annotation


We'll be using 5 broad celltypes. These are from the Garrido-Trigo et al's original paper. 

```{r}
plotUMAP(sfe, colour_by='celltype_subset', scattermore=1)
```

They're nicely distinguished on the tissue (one of the healthy control samples.)

A note on the tissue: Here the top would be the lumen of the colon, with the external epithelia at the top, and a stromal layer at the bottom. The oval-shaped epithelial layers are crypts, and that region is called the lamina propira. See: https://www.pathologyoutlines.com/topic/colonhistology.html


```{r}
plotSpatialFeature(sfe.sample.HC,'celltype_subset', colGeometryName = "cellSeg") + 
  theme(legend.title=element_blank()) +
  ggtitle(sample)
```

But there are other levels of celltype annotation in this dataset.

There is the very (very) detailed celltype_singleR2, usef for various analyses in the original paper;

```{r}
# Remove celltypes with less than 30 instances, purely for plotting.
cell_counts <- table(sfe.sample.HC$celltype_SingleR2)
sfe.sample.HC$filtered_celltype_singleR2 <- as.character(ifelse(cell_counts[as.character(sfe.sample.HC$celltype_SingleR2)] >= 30 , as.character(sfe.sample.HC$celltype_SingleR2), "Other"))

plotSpatialFeature(sfe.sample.HC, 'filtered_celltype_singleR2', colGeometryName = "cellSeg") + 
  theme(legend.title=element_blank()) +
  ggtitle(sample)
```



And some clusters genrated purely on transcriptional similarity. These were generated during preprocessing (and havnen't been translated to celltypes), but might represent a nice level of classification if were were doing the analysis from scratch. 
```{r}
plotSpatialFeature(sfe.sample.HC, 'cluster_code', colGeometryName = "cellSeg") + 
  theme(legend.title=element_blank()) +
  ggtitle(sample)
```


Of course, we can plot any gene's expression (so long as its present on the 1k panel!)
```{r}
plotSpatialFeature(sfe.sample.HC, 'PIGR', colGeometryName = "cellSeg") + 
  theme(legend.title=element_blank()) +
  ggtitle(sample)
```
