---
title: "Celltype Co-ocurance"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# This code will run the setup chunk if this is built automatically, f
# it is for building reference htmls.
# When running interactively (via Rstudio) it will not run (not needed)
# This chunk may be ignored during workshops or when running through material.
if (! base::interactive()) { source("00_Setup.R") }
```


**Are myeloid cells migrating closer to the epithelia in crohn's disease?**


We can use spicyR to test if there is a change in the relative closeness of different celltypes (e.g. myeloid and epithelia) across different condition.

See: [spicyR vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/spicyR/inst/doc/spicyR.html#41_Test_for_change_in_localisation_for_a_specific_pair_of_cells)



# Build the input table for spicyR


SpicyR can work with a SFE object, but not in this specific case (It lacks image data, and some coordinates are negative). So instead, we will build a table of the relevant input data.

For every cell, it requires the following

* Cell type
* X and Y coordinates
* Sample ID 
* Sample condition 
* FOV (optional)


The centroid coordinates of each cell are available with the spatialCoords function

```{r}
head(spatialCoords(sfe))
```

Everything else we need is stored in the ColData(). We can join those two tables together, and then rename some columns according to what spicyR expects.

```{r}
#epi => myeloid
spicy_cell_info <- as.data.frame(colData(sfe))[,c('cell', 'tissue_sample',  'celltype_subset', 'group', 'fov_name') ]
spicy_cell_info <- cbind(spicy_cell_info, spatialCoords(sfe))
spicy_cell_info <- spicy_cell_info %>%
  dplyr::rename(
    imageID = fov_name, #there are 4 fovs per tissue sample. In xenium data, that would be one per slide. And in other experiments, many samples per image.
    x=CenterX_global_px,
    y=CenterY_global_px
  )
spicy_cell_info$imageID <- droplevels(spicy_cell_info$imageID)

```


Offset any negative X/Y coordinates to positive. The absolute positions don't matter. 

```{r}
# remove negative values - not supported by spicyR
# Else yields error: Error in owinInternalRect(xrange, yrange, ..., unitname = unitname): xrange should be a vector of length 2 giving (xmin, xmax)
min_x <- min(spicy_cell_info$x)
min_y <- min(spicy_cell_info$y)
if (min_x < 0) {spicy_cell_info$x = spicy_cell_info$x + -(min_x) + 1 }
if (min_y < 0) {spicy_cell_info$y = spicy_cell_info$y + -(min_y) + 1 }
```

## Test proximity of myeloid cells to epithelia

```{r}
spicy_res <- spicy(
  cells     = spicy_cell_info , 
  subject   = 'tissue_sample',
  cellType  = 'celltype_subset',
  sigma     = 50, # tissue is not homogeneous see: https://www.bioconductor.org/packages/release/workflows/vignettes/spicyWorkflow/inst/doc/spicyWorkflow.html#14_spicyR:_Test_spatial_relationships
  condition = "group",
  from      = "epi",
  to        = "myeloids"
)
```


NB: Garrido-Trigo et al did a detailed exploration of localisation of macrophage and neutrophil subtypes in their data. We are testing a toy subset with some very broad celltypes!


```{r}
topPairs(spicy_res)
```

We can plot the 'L-function' between these groups per condition. There is an L value for each sample, which represents the 'closeness' of the celltype pair in that sample. Higher values indicates co-location. 

```{r}
spicyBoxPlot(spicy_res, 
             from = "epi", 
             to = "myeloids")
```


## Tests for changes across all celltypes


If we omit the pair of celltypes - all combinations will be tested.
```{r}
spicy_res <- spicy(
  cells     = spicy_cell_info , 
  subject   = 'tissue_sample',
  cellType  = 'celltype_subset',
  sigma     = 50, # tissue is not homogeneous see: https://www.bioconductor.org/packages/release/workflows/vignettes/spicyWorkflow/inst/doc/spicyWorkflow.html#14_spicyR:_Test_spatial_relationships
  condition = "group"
)

```

And there is a very nice visualisation. The size of the circle scales with significant, and the colour indicates the magnitude of the 'L' Value. 

```{r}
signifPlot(spicy_res)
```


We should interpret this carefully (especially for such a small dataset) - is it actually interesting biology?

```{r}
p1 <- plotSpatialFeature(sfe.sample.HC, "celltype_subset",colGeometryName = "cellSeg") + 
  theme(legend.title=element_blank()) +
  ggtitle(sample)

p2 <- plotSpatialFeature(sfe.sample.CD,  "celltype_subset", colGeometryName = "cellSeg") + 
  theme(legend.title=element_blank()) +
  ggtitle(sample)

p1 / p2 

```

